working title Compatibility system and stigma size are the main predictors of heterospecific pollen effect

### 2019-02-21 Manuscript_draft/Manuscript_draft.Rmd
edits intro

Plant traits are crucial to understand heterospecific pollen effect but the multifactorial nature of the traits that are involve in the pollen-pistil interaction make difficult to unravel what are the main traits in driving the effect. These traits can be seen from a male perspective of both donor and recipient where pollen size, pollen aperture number and pollen allelopathy are key components to understand the outcome of foreign pollen arrival @murphy1995; @@ashman2013. In @ashman2013 small pollen is predicted to cause a greater fitness decrease, although this can be true there are also other possibilities to consider which can obscure a predictive framework like big pollen can clogg small stigmas with fewer pollen grains, bigger stigmas are less likely to be clogged by small pollen grains and bigger pollen can outcompete smaller pollen grains due to faster pollen tube growth rate @williams1990. Moreover, to understand the different mechanical or chemical effects of pollen also the female traits of the pollen recipient have to be considered, from the literature these main traits are: stigma size, style length, number of ovules, incompatibility system and also a structural trait such as flower morphology @montgomery2012; @ashman2013; @tong2016. For example, greater stigmatic area is positively correlated with greater heterospecific pollen deposition @montgomery2012 and therefore possibly with an increase in negative effect. For species that are self-incompatible the barriers towards heterospecific pollen are stronger than self-compatible species @ashman2013. Nonetheless, an effect of foreign pollen is a bit obscured by the variability within species, however species that are strong selfers or strong outcrossers have less variablity in mating systems and predictions of effect could be more realistic (see figure 1 from @whitehead2018). Although past research has progress in the understanding of what traits can mediate the effect as we have shown here, there are multiple traits involved and multiple possible scenarios still to be explored empirically for a full understanding of the importance of heterospecific pollen effect in nature.

For the understanding at what level or intensity the interference of pollen can occur is important to consider the relatedness of the interacting species. Closely related species are more likely to have similar traits @arceo2016. The similarity in traits between closely related species can lead to higher chances of ovule usurpation/abortion (REF) and therefore studies predict and show a greater negative effect of closely related species (REFS). Few studies however, have focused on the impacts of heterospecific pollen on fitness of distantly related species Thomson et al. (1982) Galen and Gregory (1989) Neiland and Wilcock (1999). Yet, most insects and most stigmas have been found to carry multiple species of foreign pollen with little attention to degree of relatedness (Arceo-Gómez and Ashman (2016); Fang and Huang (2013). The relative amounts of closely/far related pollen that arrives to a stigma gives a first snapshoot where pollen competition can occur and improves the understanding of how important the reduction of fitness can be from heterospecific pollen where closely related species cand lead to greater detrimental effects @arceo2016.

>>>>

Plant traits are crucial to understand heterospecific pollen effect but the multifactorial nature of the traits that are involve in the pollen-pistil interaction make difficult to unravel what are the main traits in driving the effect. These traits can be seen from a male perspective of both donor and recipient where pollen size, pollen aperture number and pollen allelopathy are key components to understand the outcome of foreign pollen arrival @murphy1995; @ashman2013. In @ashman2013 small pollen is predicted to cause a greater fitness decrease, although this can be true there are also other possibilities to consider which can obscure a predictive framework like big pollen can clogg small stigmas with fewer pollen grains, bigger stigmas are less likely to be clogged by small pollen grains and bigger pollen can outcompete smaller pollen grains due to faster pollen tube growth rate @williams1990. Moreover, to understand the different mechanical or chemical effects of pollen also the female traits of the pollen recipient have to be considered, from the literature these main traits are: stigma size, style length, number of ovules, incompatibility system and also a structural trait such as flower morphology @montgomery2012; @ashman2013; @tong2016. For example, greater stigmatic area is positively correlated with greater heterospecific pollen deposition @montgomery2012 and therefore possibly with an increase in negative effect. For species that are self-incompatible the barriers towards heterospecific pollen are stronger than self-compatible species @ashman2013. Nonetheless, an effect of foreign pollen is a bit obscured by the variability within species, however species that are strong selfers or strong outcrossers have less variablity in mating systems and predictions of effect could be more realistic (see figure 1 from @whitehead2018). Although past research has progress in the understanding of what traits can mediate the effect as we have shown here, there are multiple traits involved and multiple possible scenarios still to be explored empirically for a full understanding of the importance of heterospecific pollen effect in nature.

For the understanding at what level or intensity the interference of pollen can occur is important to consider the relatedness of the interacting species. Closely related species are more likely to have similar traits @letten2015. The similarity in traits between closely related species can lead to higher chances of ovule usurpation/abortion (REF) and therefore studies predict and show a greater negative effect of closely related species (REFS). Few studies however, have focused on the impacts of heterospecific pollen on fitness of distantly related species Thomson et al. (1982) Galen and Gregory (1989) Neiland and Wilcock (1999). Yet, most insects and most stigmas have been found to carry multiple species of foreign pollen with little attention to degree of relatedness (Arceo-Gómez and Ashman (2016); Fang and Huang (2013). The relative amounts of closely/far related pollen that arrives to a stigma gives a first snapshoot where pollen competition can occur and improves the understanding of how important the reduction of fitness can be from heterospecific pollen where closely related species cand lead to greater detrimental effects @arceo2016.

### 2019-02-21 Manuscript_draft/Manuscript_draft.tex
edits intro

Plant traits are crucial to understand heterospecific pollen effect but
the multifactorial nature of the traits that are involve in the
pollen-pistil interaction make difficult to unravel what are the main
traits in driving the effect. These traits can be seen from a male
perspective of both donor and recipient where pollen size, pollen
aperture number and pollen allelopathy are key components to understand
the outcome of foreign pollen arrival Murphy and Aarssen (1995); @Ashman
and Arceo-Gómez (2013). In Ashman and Arceo-Gómez (2013) small pollen is
predicted to cause a greater fitness decrease, although this can be true
there are also other possibilities to consider which can obscure a
predictive framework like big pollen can clogg small stigmas with fewer
pollen grains, bigger stigmas are less likely to be clogged by small
pollen grains and bigger pollen can outcompete smaller pollen grains due
to faster pollen tube growth rate Williams and Rouse (1990). Moreover,
to understand the different mechanical or chemical effects of pollen
also the female traits of the pollen recipient have to be considered,
from the literature these main traits are: stigma size, style length,
number of ovules, incompatibility system and also a structural trait
such as flower morphology Montgomery and Rathcke (2012); Ashman and
Arceo-Gómez (2013); Tong and Huang (2016). For example, greater
stigmatic area is positively correlated with greater heterospecific
pollen deposition Montgomery and Rathcke (2012) and therefore possibly
with an increase in negative effect. For species that are
self-incompatible the barriers towards heterospecific pollen are
stronger than self-compatible species Ashman and Arceo-Gómez (2013).
Nonetheless, an effect of foreign pollen is a bit obscured by the
variability within species, however species that are strong selfers or
strong outcrossers have less variablity in mating systems and
predictions of effect could be more realistic (see figure 1 from
Whitehead et al. (2018)). Although past research has progress in the
understanding of what traits can mediate the effect as we have shown
here, there are multiple traits involved and multiple possible scenarios
still to be explored empirically for a full understanding of the
importance of heterospecific pollen effect in nature.

For the understanding at what level or intensity the interference of
pollen can occur is important to consider the relatedness of the
interacting species. Closely related species are more likely to have
similar traits Arceo-Gómez and Ashman (2016). The similarity in traits
between closely related species can lead to higher chances of ovule
usurpation/abortion (REF) and therefore studies predict and show a
greater negative effect of closely related species (REFS). Few studies
however, have focused on the impacts of heterospecific pollen on fitness
of distantly related species Thomson et al. (1982) Galen and Gregory
(1989) Neiland and Wilcock (1999). Yet, most insects and most stigmas
have been found to carry multiple species of foreign pollen with little
attention to degree of relatedness (Arceo-Gómez and Ashman (2016); Fang
and Huang (2013). The relative amounts of closely/far related pollen
that arrives to a stigma gives a first snapshoot where pollen
competition can occur and improves the understanding of how important
the reduction of fitness can be from heterospecific pollen where closely
related species cand lead to greater detrimental effects Arceo-Gómez and
Ashman (2016).

@@

>>>>

Plant traits are crucial to understand heterospecific pollen effect but
the multifactorial nature of the traits that are involve in the
pollen-pistil interaction make difficult to unravel what are the main
traits in driving the effect. These traits can be seen from a male
perspective of both donor and recipient where pollen size, pollen
aperture number and pollen allelopathy are key components to understand
the outcome of foreign pollen arrival Murphy and Aarssen (1995); Ashman
and Arceo-Gómez (2013). In Ashman and Arceo-Gómez (2013) small pollen is
predicted to cause a greater fitness decrease, although this can be true
there are also other possibilities to consider which can obscure a
predictive framework like big pollen can clogg small stigmas with fewer
pollen grains, bigger stigmas are less likely to be clogged by small
pollen grains and bigger pollen can outcompete smaller pollen grains due
to faster pollen tube growth rate Williams and Rouse (1990). Moreover,
to understand the different mechanical or chemical effects of pollen
also the female traits of the pollen recipient have to be considered,
from the literature these main traits are: stigma size, style length,
number of ovules, incompatibility system and also a structural trait
such as flower morphology Montgomery and Rathcke (2012); Ashman and
Arceo-Gómez (2013); Tong and Huang (2016). For example, greater
stigmatic area is positively correlated with greater heterospecific
pollen deposition Montgomery and Rathcke (2012) and therefore possibly
with an increase in negative effect. For species that are
self-incompatible the barriers towards heterospecific pollen are
stronger than self-compatible species Ashman and Arceo-Gómez (2013).
Nonetheless, an effect of foreign pollen is a bit obscured by the
variability within species, however species that are strong selfers or
strong outcrossers have less variablity in mating systems and
predictions of effect could be more realistic (see figure 1 from
Whitehead et al. (2018)). Although past research has progress in the
understanding of what traits can mediate the effect as we have shown
here, there are multiple traits involved and multiple possible scenarios
still to be explored empirically for a full understanding of the
importance of heterospecific pollen effect in nature.

For the understanding at what level or intensity the interference of
pollen can occur is important to consider the relatedness of the
interacting species. Closely related species are more likely to have
similar traits Letten and Cornwell (2015). The similarity in traits
between closely related species can lead to higher chances of ovule
usurpation/abortion (REF) and therefore studies predict and show a
greater negative effect of closely related species (REFS). Few studies
however, have focused on the impacts of heterospecific pollen on fitness
of distantly related species Thomson et al. (1982) Galen and Gregory
(1989) Neiland and Wilcock (1999). Yet, most insects and most stigmas
have been found to carry multiple species of foreign pollen with little
attention to degree of relatedness (Arceo-Gómez and Ashman (2016); Fang
and Huang (2013). The relative amounts of closely/far related pollen
that arrives to a stigma gives a first snapshoot where pollen
competition can occur and improves the understanding of how important
the reduction of fitness can be from heterospecific pollen where closely
related species cand lead to greater detrimental effects Arceo-Gómez and
Ashman (2016).

@@

\hypertarget{ref-letten2015}{}
Letten, A. D., and W. K. Cornwell. 2015. Trees, branches and (square)
roots: Why evolutionary relatedness is not linearly related to
functional distance. Methods in Ecology and Evolution 6:439--444.
